Asteroid project summary

### 2022-01-14 summaries/21-11/21-11.tex
Probe space data compiled

\usepackage{tikz}
\usepackage{dsfont}

\newcommand{\unit}[1]{\hat{\mathbf{#1}}}
\newcommand{\parens}[1]{\left( #1 \right)}
\newcommand{\brackets}[1]{\left[ #1 \right]}
\newcommand{\jtd}[1]{{\color{red}\textit{#1}}}
\newcommand{\quat}[1]{\widetilde{\mathbf{#1}}}
\newcommand{\siunit}[1]{\ \textrm{#1}}
\newcommand{\per}[1]{\ \textrm{#1}^{-1}}
\usetikzlibrary{arrows.meta}

@@

\begin{itemize}
\item Asteroids simulated in binaries \cite{Naidu_2015}
\item Previous work on the math of asteroid rotations %\cite{ashenberg07, paul88}
\item CNEOS: 4 asteroids $D > 100 m$ are coming to within 1 LD of Earth in the next 10 years, 29 with $D > 10 m$.
\end{itemize}

@@

We define a rotation matrix $M$ such that $M\mathbf{r} = \mathbf{R}$, where $\mathbf{r}$ is in the body-fixed frame and $\mathbf{R}$ is in the inertial frame. It will be useful for us to represent this matrix in two different ways; one is with a normalized quaternion $\quat q$, and the other is with $z-y-z$ Euler angles named $\alpha$, $\beta$, and $\gamma$. In particular, we define
\begin{equation}
M = R_z(\alpha) R_y(\beta) R_z(\gamma)
\label{eqn:euler-angles}
\end{equation}
where $R_i(\theta)$ is a rotation around the $i$th unit vector by $\theta$ (figure \ref{fig:euler-angles}).

@@

We use the quaternion $\quat q$ as our dynamical variable because Euler angles suffer from gimbal lock. For the sake of this paper, we write quaternions as $\quat q = r + i \mathbf i + j \mathbf j + k \mathbf k$, with the complex conjugate written as $\quat q^* = r - i \mathbf i - j \mathbf j - k \mathbf k$. The inverse is then $\quat q^{-1} = \quat q^* / (\quat q \quat q^*)$, and we define real and imaginary parts as $\Re \quat q = r$; $\Im \quat q = i \unit x + j \unit y + k \unit z$. See a description of quaternions in Ref.~\cite{2008arXiv0811.2889G}

Normalized quaternions express rotations of an angle $\theta$ by $\quat q = \cos\frac{\theta}{2} + \sin\frac{\theta}{2}\quat n$ where $\quat n = 0 + n_x \mathbf i + n_y \mathbf j + n_z\mathbf k$. This rotation may be applied to any vector $\mathbf r$ by
\begin{equation}
\mathbf R = \Im \brackets{\quat q \quat r\quat q^{-1} }.
\label{eqn:quat-rot}
\end{equation}
Due to the real part of $\quat q$ which does not vanish for $\theta \rightarrow 0$, quaternions do not suffer from gimbal lock. Using the above notation, we can give the full equation of quaternion multiplication:
\begin{equation}
(q_0 + \quat q) (p_0 + \quat p) = q_0 p_0 - \mathbf p \cdot \mathbf q + p_0 \mathbf q + q_0 \mathbf p + \mathbf q \times \mathbf p.
\label{eqn:quat-mult}
\end{equation}
Note that quaternion multiplication is not commutative.

@@

\subsection{Tidal Torque}
We start by expressing the gravitational potential energy field of the central body in its most general form:
\begin{equation}
V(\mathbf R') = -G\int_\mathcal{P} d^3 R \rho_M(\mathbf R) \frac{1}{|\mathbf{R}-\mathbf{R'}|}.
\label{eqn:first-pe}
\end{equation}
This integral is computed in the inertial frame and has units of energy per unit mass. We represent the central body density as $\rho_M$ and its volume as $\mathcal{P}$. To reduce equation \ref{eqn:first-pe}, we define the unnormalized spherical harmonics $Y_{\ell m}(\theta, \phi) = P_{\ell m}(\cos \theta)e^{im\phi}$, where $P_{\ell m}$ are the associated Legendre Polynomials without the Condon-Shortley phase. The regular and irregular spherical harmonics are then defined as
\begin{equation}
S_{\ell m}(\mathbf r) = (-1)^m (\ell - m)! \frac{Y_{\ell m}(\unit r)}{r^{\ell+1}} \qquad R_{\ell m} (\mathbf r) = (-1)^m \frac{r^\ell}{(\ell + m)!} Y_{\ell m}(\unit r).
\end{equation}

Assume that we can construct a sphere, centered at the center of mass of the central body, so that $\rho_M = 0$ outside the sphere and $\rho_m = 0$ inside the sphere. Then $R' > R$ in equation \ref{eqn:first-pe}, and Ref.~\cite{Gelderen1998TheSO} gives the identity
\begin{equation}
\frac{1}{|\mathbf R - \mathbf R'|} = \sum_{\ell, m} R_{\ell m}(\mathbf R) S_{\ell m}^*(\mathbf R'),
\label{eqn:1-over-r-expansion}
\end{equation}
where the sum is shorthand for $\sum_{\ell, m} = \sum_{\ell = 0}^\infty \sum_{m=-\ell}^\ell$.

Suppose the vector $\mathbf{D}$ points from the center of mass of the central body to the center of mass of the asteroid, again in the inertial frame. Then we may write $ \mathbf R' = \mathbf D + \mathbf U$, and Ref.~\cite{Gelderen1998TheSO} gives
\begin{equation}
S_{\ell m}(\mathbf R') = \sum_{\ell', m'} (-1)^{\ell'}R^*_{\ell' m'}(\mathbf U)S_{\ell+\ell', m + m'} (\mathbf D).
\end{equation}

Define the complex, unitless constants
\begin{equation}
J_{\ell m} = \int_\mathcal{P} d^3 R \rho_M(\mathbf R) \frac{R_{\ell m}(\mathbf R)}{\mu_M a_M^\ell}
\qquad
K_{\ell m} = \int_\mathcal{A} d^3 r \rho_m(\mathbf r) \frac{R_{\ell m}(\mathbf r)}{\mu_m a_m^\ell}
\label{eqn:jlm}
\end{equation}
where the integrals are taken in the inertial and the body-fixed frames respectively. Here, $\rho_m$ denotes the density of the asteroid and $\mathcal{A}$ denotes its volume. The symmetry relations of $R_{\ell m}$ guarantee $J_{\ell m} = (-1)^m J_{\ell, -m}^*$ and similarly for $K_{\ell m}$. We have chosen $\mu_M$, $\mu_m$ to be the masses of the central body and asteroid respectively, and $a_M$ and $a_m$ to be scale lengths of both bodies (defined more precisely later). Then we can express the potential as
\begin{equation}
V(\mathbf U) = -G \brackets{\sum_{\ell, m} \mu_M a_M^\ell J_{\ell m} \sum_{\ell', m'} (-1)^{\ell'}R_{\ell' m'}(\mathbf U)S^*_{\ell+\ell', m + m'} (\mathbf D)}.
\label{eqn:pe-unrotated}
\end{equation}
To make use of $K_{\ell m}$ in equation \ref{eqn:jlm}, we need to rotate $\mathbf U$ from the inertial frame into $\mathbf u$ in the body fixed frame via $\mathbf U = M \mathbf u$. The rotation of a spherical harmonic satisfies
\begin{equation}
Y_{\ell m}(M\mathbf u) = \sum_{m'=-\ell}^\ell (-1)^{m+m'}\sqrt{\frac{(\ell-m')!(\ell+m)!}{(\ell+m')!(\ell-m)!}} \mathcal{D}^\ell_{mm'}(M)^* Y_{\ell m'}(\mathbf u).
\end{equation}
where $\mathcal{D}$ are the Wigner-$D$ matrices. The corollary of this formula for the regular solid spherical harmonic is
\begin{equation}
R_{\ell m} (M \mathbf u) = \sum_{m'=-\ell}^\ell \sqrt{\frac{(\ell-m')!(\ell+m')!}{(\ell-m)!(\ell+m)!}} \mathcal{D}^\ell_{mm'}(M)^* R_{\ell m'} (\mathbf u)
\label{eqn:rlm-rotations}
\end{equation}
where $\mathcal{D}^\ell_{mm'}$ are the Wigner-D matrix elements. In practice, these are computed from the Euler angles corresponding to the rotation $M$, which we specified in section \ref{sec:coordinates} are $z-y-z$ Euler angles. With this choice of Euler angles, the Wigner-D matrix elements are real with matrix elements given by
\begin{eqnarray}
d^\ell_{mm'}(\beta) = &\sqrt{(\ell+m)!(\ell-m)!(\ell+m')!(\ell-m')!}\\
& \sum_{s=s_\textrm{min}}^{s_\textrm{max}} \brackets{\frac{(-1)^{m-m'+s}\parens{\cos\frac{\beta}{2}}^{2\ell + m' - m - 2s}\parens{\sin \frac{\beta}{2}}^{m-m'+2s}}{(\ell+m'-s)!(\ell-m-s)!(m-m'+s)!s!}}
\label{eqn:wigner-d-elements}
\end{eqnarray}
where $s_\textrm{min} = \max(0, m'-m)$ and $s_\textrm{max} = \min(\ell+m', \ell-m)$,
and $\mathcal{D}^\ell_{mm'}(\alpha, \beta, \gamma) = d^j_{mm'}e^{-im\alpha} e^{-im'\gamma}$.

Combining equations \ref{eqn:rlm-rotations} and \ref{eqn:pe-unrotated},
\begin{equation}
V(\mathbf u) = -G \brackets{\sum_{\ell, m} \mu_M a_M^\ell J_{\ell m} \sum_{\ell', m'} (-1)^{\ell'}S^*_{\ell+\ell', m + m'} (\mathbf D)\sum_{m''=-\ell'}^{\ell'} \sqrt{\frac{(\ell'-m'')!(\ell'+m'')!}{(\ell'-m')!(\ell'+m')!}} \mathcal{D}^{\ell'}_{m'm''}(M)^* R_{\ell' m''} (\mathbf u)}.
\end{equation}

The tidal torque applied to an asteroid at a point $\mathbf u$ is given by $d\mathbf \tau(\mathbf u) = -\rho_m(\mathbf u) d^3u (\mathbf u \times (\nabla V_{\mathbf u}(\mathbf u)))$ in the body-fixed frame, which leaves
\begin{eqnarray}
\mathbf\tau = &G\int_\mathcal{A}d^3u \rho_m(\mathbf u)  \mathbf u \times \nabla \left[\sum_{\ell, m} \mu_M a_M^\ell J_{\ell m} \sum_{\ell', m'} (-1)^{\ell'}S^*_{\ell+\ell', m + m'} (\mathbf D)
\right.\\
& \left. \sum_{m''=-\ell'}^{\ell'} \sqrt{\frac{(\ell'-m'')!(\ell'+m'')!}{(\ell'-m')!(\ell'-m')!}} \mathcal{D}^{\ell'}_{m'm''}(M)^* R_{\ell' m''} (\mathbf u)\right].
\label{eqn:first-tidal-torque}
\end{eqnarray}
Using the identity
\begin{eqnarray}
\mathbf u \times \nabla R_{\ell m}(\mathbf u)=&\frac{1}{2}[(i\unit x - \unit y)(\ell-m+1)R_{\ell,m-1}(\mathbf u)\\
&+(i\unit x+\unit y)(\ell+m+1)R_{\ell,m+1}(\mathbf u)+2im\unit z R_{\ell m}(\mathbf u)],
\end{eqnarray}
which was derived from Ref.~\cite{Gelderen1998TheSO}'s analysis of the angular momentum operators applied to spherical harmonics, we write equation \ref{eqn:first-tidal-torque} as
\begin{eqnarray}
\mathbf \tau = &G\frac{\mu_m\mu_M}{2}\left[\sum_{\ell, m} a_M^\ell J_{\ell m} \sum_{\ell', m'}a_m^{\ell'}S^*_{\ell+\ell', m + m'} (\mathbf D) (-1)^{\ell'}\sum_{m''=-\ell'}^{\ell'} \sqrt{\frac{(\ell'-m'')!(\ell'+m'')!}{(\ell'-m')!(\ell'+m')!}}  \mathcal{D}^{\ell'}_{m'm''}(M)^* \right. \\
&\bigg[(i\unit x - \unit y)(\ell'-m''+1)K_{\ell',m''-1} +(i\unit x+\unit y)(\ell'+m''+1)K_{\ell',m''+1}+2im''\unit z K_{\ell'm''}\bigg] \Bigg].
\label{eqn:tidal-torque}
\end{eqnarray}
This equation contains complex values, but it was numerically confirmed that $\mathbf \tau$ is a real-valued vector.

@@

\subsection{Moment of Inertia}
The inertia tensor of the asteroid is expressed as a matrix $I$. Written in terms of spherical harmonics in the body-fixed frame, we have
\begin{eqnarray}
I & = &\int_\mathcal{A} d^3 r\rho_m(\mathbf r) \parens{\begin{array}{ccc}
y^2 + z^2 & -xy & -xz \\
-xy & x^2 + z^2 & -xz \\
-xz & -yz & x^2 + y^2 \\
\end{array}}\\
& = &\frac{2}{3}\mu_m a_m^2\parens{\begin{array}{ccc}
K_{20}-3K_{2,-2}-3K_{22}+1 & 3i \parens{K_{22}-K_{2,-2}} & \frac{3}{2}\parens{K_{21}-K_{2,-1}}\\
3i \parens{K_{22}-K_{2,-2}} & K_{20}+3K_{2,-2}+3K_{22}+1 & -\frac{3}{2}i\parens{K_{21}+K_{2,-1}}\\
\frac{3}{2}\parens{K_{21}-K_{2,-1}} & -\frac{3}{2}i\parens{K_{21}+K_{2,-1}} & -2K_{20}+1\\
\end{array}}
\label{eqn:moi}
\end{eqnarray}
where we have defined, for neatness,
\begin{eqnarray}
a_m^2=\frac{1}{\mu_m}\int_\mathcal{A} d^3 r \rho_m(\mathbf r) r^2.
\label{eqn:am}
\end{eqnarray}
With this definition of $a_m$, and an arbitrary definition of $a_M$ which may be taken to be similar to equation \ref{eqn:am} or a more convenient unit such as the central body radius, the parameters $J_{\ell m}$ and $K_{\ell m}$ of equation \ref{eqn:jlm} are completely defined. Note that $a_m$ is similar to the radius of the asteroid; if the asteroid is a sphere of uniform density and radius $r$, then $a_m = r\sqrt{3/5}$.

@@

\subsection{Equation of Motion}
As a simplifying assumption, we will not model the effect of the central body's non-sphericity on the asteroid orbit. This effect has been studied in much detail and is small. We therefore separate the position dynamical variables from the angular variables and derive the standard equations of motion for asteroid position in the inertial frame:
\begin{equation}
\dot{\mathbf{V}} = -\frac{G\mu_M}{r^3}\mathbf{D}\qquad \dot{\mathbf{D}} = \mathbf{V}.
\label{eqn:translaton-eom}
\end{equation}
The position variables will no longer be discussed.

Let $\mathbf\omega$ be the rotational velocity of the asteroid in the body-fixed frame. We have two equations of motion for our system; one governs the evolution of $\mathbf \omega$ and the other governs the evolution of $\mathbf q$. Starting with the latter, consider a fixed point $\mathbf r$ in the body-fixed frame on a rotating asteroid. The corresponding point $\mathbf R=\Im[\quat q \mathbf r \quat q^{-1}]$ in the inertial frame satisfies $\dot{\mathbf R} = \Im[\quat q(\mathbf \omega \times \mathbf r )\quat q^{-1}]$. Furthermore, we can demand that $\mathbf \omega \perp \mathbf r$ so that $\mathbf \omega \times \mathbf r = \Im[\quat \omega \quat r]$. We obtain another equation for $\dot{\mathbf R}$ by differentiating equation \ref{eqn:quat-rot} with $\dot{\mathbf r} = 0$. Setting these equal,
\begin{equation}
\dot{\quat q} \quat r \quat q^{-1} + \quat q \quat r \dot{\quat q}^{-1} = \quat q\quat \omega \quat r \quat q^{-1}.
\nonumber
\end{equation}
Substituting $\mathbf r \rightarrow \quat q^{-1}\mathbf R \quat q$ and acknowledging that $\Re[\dot{\quat q} \quat q^{-1}] = 0$,
\begin{equation}
  \dot{\quat q} \quat q^{-1}\mathbf R + \mathbf R \quat q \dot{\quat q}^{-1} = 2\dot{\quat q}\quat q^{-1}\quat R = \quat q\quat \omega \quat q^{-1}\mathbf R.
\nonumber
\end{equation}
Since $\mathbf R$ is arbitrary, this simplifies to
\begin{equation}
  2\dot{\quat q}\quat q^{-1} =  \quat q\quat \omega \quat q^{-1} \Rightarrow \dot{\quat q} = \frac{1}{2}\quat q\quat \omega.
\label{eqn:quat-eom}
\end{equation}

The second equation of motion are Euler's equations of motion given our expression for tidal torque. Namely,
\begin{eqnarray}
I_1 \dot \omega_1 - \omega_2 \omega_3 (I_2 - I_3) &= \tau_1\\
I_2 \dot \omega_2 - \omega_3 \omega_1 (I_3 - I_1) &= \tau_2\\
I_3 \dot \omega_3 - \omega_1 \omega_2 (I_1 - I_2) &= \tau_3
\label{eqn:omega-eom}
\end{eqnarray}
The components of $\mathbf \tau$ are given by equation \ref{eqn:tidal-torque}, which was also in the body-fixed frame.

\subsection{Initial Values}
We assume that the asteroid begins the simulation in its minimum energy configuration; that is, with $\mathbf \omega \parallel \unit z$. The initial conditions are therefore that $\mathbf \omega = \omega \unit z$, and the quaternion is fixed to rotate $\hat z$ to $\mathbf \Omega$ in the inertial frame, which fixes Euler angles $\beta$ and $\alpha$. The third Euler angle $\gamma$ represents the initial roll of the asteroid, and therefore cannot be determined from physical data. However, it is an initial condition of the simulation. Once the Euler angles are known, the quaternion is computed via equation \ref{eqn:euler-to-quat}.

@@

\jtd{Running Jan 11?}

@@

\bibliography{asteroids}{}
\bibliographystyle{aasjournal}

>>>>

\usepackage{tikz}
\usepackage{dsfont}
\usepackage{bm}

\newcommand{\unit}[1]{\hat{\bm{#1}}}
\newcommand{\parens}[1]{\left( #1 \right)}
\newcommand{\brackets}[1]{\left[ #1 \right]}
\newcommand{\jtd}[1]{{\color{red}\textit{#1}}}
\newcommand{\quat}[1]{\widetilde{\bm{#1}}}
\newcommand{\siunit}[1]{\ \textrm{#1}}
\newcommand{\per}[1]{\ \textrm{#1}^{-1}}
\usetikzlibrary{arrows.meta}

@@

\begin{itemize}
\item Asteroids simulated in binaries \cite{Naidu_2015}
\item Previous work on the math of asteroid rotations \cite{ashenberg07, paul88}
\item CNEOS: 4 asteroids $D > 100 m$ are coming to within 1 LD of Earth in the next 10 years, 29 with $D > 10 m$.
\end{itemize}

@@

We define a rotation matrix $M$ such that $M\bm{r} = \bm{R}$, where $\bm{r}$ is in the body-fixed frame and $\bm{R}$ is in the inertial frame. It will be useful for us to represent this matrix in two different ways; one is with a normalized quaternion $\quat q$, and the other is with $z-y-z$ Euler angles named $\alpha$, $\beta$, and $\gamma$. In particular, we define
\begin{equation}
M = R_z(\alpha) R_y(\beta) R_z(\gamma)
\label{eqn:euler-angles}
\end{equation}
where $R_i(\theta)$ is a rotation around the $i$th unit vector by $\theta$ (figure \ref{fig:euler-angles}).

@@

We use the quaternion $\quat q$ as our dynamical variable because Euler angles suffer from gimbal lock. For the sake of this paper, we write quaternions as $\quat q = r + i \bm i + j \bm j + k \bm k$, with the complex conjugate written as $\quat q^* = r - i \bm i - j \bm j - k \bm k$. The inverse is then $\quat q^{-1} = \quat q^* / (\quat q \quat q^*)$, and we define real and imaginary parts as $\Re \quat q = r$; $\Im \quat q = i \unit x + j \unit y + k \unit z$. See a description of quaternions in Ref.~\cite{2008arXiv0811.2889G}

Normalized quaternions express rotations of an angle $\theta$ by $\quat q = \cos\frac{\theta}{2} + \sin\frac{\theta}{2}\quat n$ where $\quat n = 0 + n_x \bm i + n_y \bm j + n_z\bm k$. This rotation may be applied to any vector $\bm r$ by
\begin{equation}
\bm R = \Im \brackets{\quat q \quat r\quat q^{-1} }.
\label{eqn:quat-rot}
\end{equation}
Due to the real part of $\quat q$ which does not vanish for $\theta \rightarrow 0$, quaternions do not suffer from gimbal lock. Using the above notation, we can give the full equation of quaternion multiplication:
\begin{equation}
(q_0 + \quat q) (p_0 + \quat p) = q_0 p_0 - \bm p \cdot \bm q + p_0 \bm q + q_0 \bm p + \bm q \times \bm p.
\label{eqn:quat-mult}
\end{equation}
Note that quaternion multiplication is not commutative.

@@

\subsection{Tidal Torque}
We start by expressing the gravitational potential energy field of the central body in its most general form:
\begin{equation}
V(\bm R') = -G\int_\mathcal{P} d^3 R \rho_M(\bm R) \frac{1}{|\bm{R}-\bm{R'}|}.
\label{eqn:first-pe}
\end{equation}
This integral is computed in the inertial frame and has units of energy per unit mass. We represent the central body density as $\rho_M$ and its volume as $\mathcal{P}$. To reduce equation \ref{eqn:first-pe}, we define the unnormalized spherical harmonics $Y_{\ell m}(\theta, \phi) = P_{\ell m}(\cos \theta)e^{im\phi}$, where $P_{\ell m}$ are the associated Legendre Polynomials without the Condon-Shortley phase. The regular and irregular spherical harmonics are then defined as
\begin{equation}
S_{\ell m}(\bm r) = (-1)^m (\ell - m)! \frac{Y_{\ell m}(\unit r)}{r^{\ell+1}} \qquad R_{\ell m} (\bm r) = (-1)^m \frac{r^\ell}{(\ell + m)!} Y_{\ell m}(\unit r).
\end{equation}

Assume that we can construct a sphere, centered at the center of mass of the central body, so that $\rho_M = 0$ outside the sphere and $\rho_m = 0$ inside the sphere. Then $R' > R$ in equation \ref{eqn:first-pe}, and Ref.~\cite{Gelderen1998TheSO} gives the identity
\begin{equation}
\frac{1}{|\bm R - \bm R'|} = \sum_{\ell, m} R_{\ell m}(\bm R) S_{\ell m}^*(\bm R'),
\label{eqn:1-over-r-expansion}
\end{equation}
where the sum is shorthand for $\sum_{\ell, m} = \sum_{\ell = 0}^\infty \sum_{m=-\ell}^\ell$.

Suppose the vector $\bm{D}$ points from the center of mass of the central body to the center of mass of the asteroid, again in the inertial frame. Then we may write $ \bm R' = \bm D + \bm U$, and Ref.~\cite{Gelderen1998TheSO} gives
\begin{equation}
S_{\ell m}(\bm R') = \sum_{\ell', m'} (-1)^{\ell'}R^*_{\ell' m'}(\bm U)S_{\ell+\ell', m + m'} (\bm D).
\end{equation}

Define the complex, unitless constants
\begin{equation}
J_{\ell m} = \int_\mathcal{P} d^3 R \rho_M(\bm R) \frac{R_{\ell m}(\bm R)}{\mu_M a_M^\ell}
\qquad
K_{\ell m} = \int_\mathcal{A} d^3 r \rho_m(\bm r) \frac{R_{\ell m}(\bm r)}{\mu_m a_m^\ell}
\label{eqn:jlm}
\end{equation}
where the integrals are taken in the inertial and the body-fixed frames respectively. Here, $\rho_m$ denotes the density of the asteroid and $\mathcal{A}$ denotes its volume. The symmetry relations of $R_{\ell m}$ guarantee $J_{\ell m} = (-1)^m J_{\ell, -m}^*$ and similarly for $K_{\ell m}$. We have chosen $\mu_M$, $\mu_m$ to be the masses of the central body and asteroid respectively, and $a_M$ and $a_m$ to be scale lengths of both bodies (defined more precisely later). Then we can express the potential as
\begin{equation}
V(\bm U) = -G \brackets{\sum_{\ell, m} \mu_M a_M^\ell J_{\ell m} \sum_{\ell', m'} (-1)^{\ell'}R_{\ell' m'}(\bm U)S^*_{\ell+\ell', m + m'} (\bm D)}.
\label{eqn:pe-unrotated}
\end{equation}
To make use of $K_{\ell m}$ in equation \ref{eqn:jlm}, we need to rotate $\bm U$ from the inertial frame into $\bm u$ in the body fixed frame via $\bm U = M \bm u$. The rotation of a spherical harmonic satisfies
\begin{equation}
Y_{\ell m}(M\bm u) = \sum_{m'=-\ell}^\ell (-1)^{m+m'}\sqrt{\frac{(\ell-m')!(\ell+m)!}{(\ell+m')!(\ell-m)!}} \mathcal{D}^\ell_{mm'}(M)^* Y_{\ell m'}(\bm u).
\end{equation}
where $\mathcal{D}$ are the Wigner-$D$ matrices. The corollary of this formula for the regular solid spherical harmonic is
\begin{equation}
R_{\ell m} (M \bm u) = \sum_{m'=-\ell}^\ell \sqrt{\frac{(\ell-m')!(\ell+m')!}{(\ell-m)!(\ell+m)!}} \mathcal{D}^\ell_{mm'}(M)^* R_{\ell m'} (\bm u)
\label{eqn:rlm-rotations}
\end{equation}
where $\mathcal{D}^\ell_{mm'}$ are the Wigner-D matrix elements. In practice, these are computed from the Euler angles corresponding to the rotation $M$, which we specified in section \ref{sec:coordinates} are $z-y-z$ Euler angles. With this choice of Euler angles, the Wigner-D matrix elements are real with matrix elements given by
\begin{eqnarray}
d^\ell_{mm'}(\beta) = &\sqrt{(\ell+m)!(\ell-m)!(\ell+m')!(\ell-m')!}\\
& \sum_{s=s_\textrm{min}}^{s_\textrm{max}} \brackets{\frac{(-1)^{m-m'+s}\parens{\cos\frac{\beta}{2}}^{2\ell + m' - m - 2s}\parens{\sin \frac{\beta}{2}}^{m-m'+2s}}{(\ell+m'-s)!(\ell-m-s)!(m-m'+s)!s!}}
\label{eqn:wigner-d-elements}
\end{eqnarray}
where $s_\textrm{min} = \max(0, m'-m)$ and $s_\textrm{max} = \min(\ell+m', \ell-m)$,
and $\mathcal{D}^\ell_{mm'}(\alpha, \beta, \gamma) = d^j_{mm'}e^{-im\alpha} e^{-im'\gamma}$.

Combining equations \ref{eqn:rlm-rotations} and \ref{eqn:pe-unrotated},
\begin{equation}
V(\bm u) = -G \brackets{\sum_{\ell, m} \mu_M a_M^\ell J_{\ell m} \sum_{\ell', m'} (-1)^{\ell'}S^*_{\ell+\ell', m + m'} (\bm D)\sum_{m''=-\ell'}^{\ell'} \sqrt{\frac{(\ell'-m'')!(\ell'+m'')!}{(\ell'-m')!(\ell'+m')!}} \mathcal{D}^{\ell'}_{m'm''}(M)^* R_{\ell' m''} (\bm u)}.
\end{equation}

The tidal torque applied to an asteroid at a point $\bm u$ is given by $d\bm \tau(\bm u) = -\rho_m(\bm u) d^3u (\bm u \times (\nabla V_{\bm u}(\bm u)))$ in the body-fixed frame, which leaves
\begin{eqnarray}
\bm\tau = &G\int_\mathcal{A}d^3u \rho_m(\bm u)  \bm u \times \nabla \left[\sum_{\ell, m} \mu_M a_M^\ell J_{\ell m} \sum_{\ell', m'} (-1)^{\ell'}S^*_{\ell+\ell', m + m'} (\bm D)
\right.\\
& \left. \sum_{m''=-\ell'}^{\ell'} \sqrt{\frac{(\ell'-m'')!(\ell'+m'')!}{(\ell'-m')!(\ell'-m')!}} \mathcal{D}^{\ell'}_{m'm''}(M)^* R_{\ell' m''} (\bm u)\right].
\label{eqn:first-tidal-torque}
\end{eqnarray}
Using the identity
\begin{eqnarray}
\bm u \times \nabla R_{\ell m}(\bm u)=&\frac{1}{2}[(i\unit x - \unit y)(\ell-m+1)R_{\ell,m-1}(\bm u)\\
&+(i\unit x+\unit y)(\ell+m+1)R_{\ell,m+1}(\bm u)+2im\unit z R_{\ell m}(\bm u)],
\end{eqnarray}
which was derived from Ref.~\cite{Gelderen1998TheSO}'s analysis of the angular momentum operators applied to spherical harmonics, we write equation \ref{eqn:first-tidal-torque} as
\begin{eqnarray}
\bm \tau = &G\frac{\mu_m\mu_M}{2}\left[\sum_{\ell, m} a_M^\ell J_{\ell m} \sum_{\ell', m'}a_m^{\ell'}S^*_{\ell+\ell', m + m'} (\bm D) (-1)^{\ell'}\sum_{m''=-\ell'}^{\ell'} \sqrt{\frac{(\ell'-m'')!(\ell'+m'')!}{(\ell'-m')!(\ell'+m')!}}  \mathcal{D}^{\ell'}_{m'm''}(M)^* \right. \\
&\bigg[(i\unit x - \unit y)(\ell'-m''+1)K_{\ell',m''-1} +(i\unit x+\unit y)(\ell'+m''+1)K_{\ell',m''+1}+2im''\unit z K_{\ell'm''}\bigg] \Bigg].
\label{eqn:tidal-torque}
\end{eqnarray}
This equation contains complex values, but it was numerically confirmed that $\bm \tau$ is a real-valued vector.

@@

\subsection{Moment of Inertia}
The inertia tensor of the asteroid is expressed as a matrix $I$. Written in terms of spherical harmonics in the body-fixed frame, we have
\begin{eqnarray}
I & = &\int_\mathcal{A} d^3 r\rho_m(\bm r) \parens{\begin{array}{ccc}
y^2 + z^2 & -xy & -xz \\
-xy & x^2 + z^2 & -xz \\
-xz & -yz & x^2 + y^2 \\
\end{array}}\\
& = &\frac{2}{3}\mu_m a_m^2\parens{\begin{array}{ccc}
K_{20}-3K_{2,-2}-3K_{22}+1 & 3i \parens{K_{22}-K_{2,-2}} & \frac{3}{2}\parens{K_{21}-K_{2,-1}}\\
3i \parens{K_{22}-K_{2,-2}} & K_{20}+3K_{2,-2}+3K_{22}+1 & -\frac{3}{2}i\parens{K_{21}+K_{2,-1}}\\
\frac{3}{2}\parens{K_{21}-K_{2,-1}} & -\frac{3}{2}i\parens{K_{21}+K_{2,-1}} & -2K_{20}+1\\
\end{array}}
\label{eqn:moi}
\end{eqnarray}
where we have defined, for neatness,
\begin{eqnarray}
a_m^2=\frac{1}{\mu_m}\int_\mathcal{A} d^3 r \rho_m(\bm r) r^2.
\label{eqn:am}
\end{eqnarray}
With this definition of $a_m$, and an arbitrary definition of $a_M$ which may be taken to be similar to equation \ref{eqn:am} or a more convenient unit such as the central body radius, the parameters $J_{\ell m}$ and $K_{\ell m}$ of equation \ref{eqn:jlm} are completely defined. Note that $a_m$ is similar to the radius of the asteroid; if the asteroid is a sphere of uniform density and radius $r$, then $a_m = r\sqrt{3/5}$.

@@

\subsection{Equation of Motion}
As a simplifying assumption, we will not model the effect of the central body's non-sphericity on the asteroid orbit. This effect has been studied in much detail and is small. We therefore separate the position dynamical variables from the angular variables and derive the standard equations of motion for asteroid position in the inertial frame:
\begin{equation}
\dot{\bm{V}} = -\frac{G\mu_M}{r^3}\bm{D}\qquad \dot{\bm{D}} = \bm{V}.
\label{eqn:translaton-eom}
\end{equation}
The position variables will no longer be discussed.

Let $\bm\omega$ be the rotational velocity of the asteroid in the body-fixed frame. We have two equations of motion for our system; one governs the evolution of $\bm \omega$ and the other governs the evolution of $\bm q$. Starting with the latter, consider a fixed point $\bm r$ in the body-fixed frame on a rotating asteroid. The corresponding point $\bm R=\Im[\quat q \bm r \quat q^{-1}]$ in the inertial frame satisfies $\dot{\bm R} = \Im[\quat q(\bm \omega \times \bm r )\quat q^{-1}]$. Furthermore, we can demand that $\bm \omega \perp \bm r$ so that $\bm \omega \times \bm r = \Im[\quat \omega \quat r]$. We obtain another equation for $\dot{\bm R}$ by differentiating equation \ref{eqn:quat-rot} with $\dot{\bm r} = 0$. Setting these equal,
\begin{equation}
\dot{\quat q} \quat r \quat q^{-1} + \quat q \quat r \dot{\quat q}^{-1} = \quat q\quat \omega \quat r \quat q^{-1}.
\nonumber
\end{equation}
Substituting $\bm r \rightarrow \quat q^{-1}\bm R \quat q$ and acknowledging that $\Re[\dot{\quat q} \quat q^{-1}] = 0$,
\begin{equation}
  \dot{\quat q} \quat q^{-1}\bm R + \bm R \quat q \dot{\quat q}^{-1} = 2\dot{\quat q}\quat q^{-1}\quat R = \quat q\quat \omega \quat q^{-1}\bm R.
\nonumber
\end{equation}
Since $\bm R$ is arbitrary, this simplifies to
\begin{equation}
  2\dot{\quat q}\quat q^{-1} =  \quat q\quat \omega \quat q^{-1} \Rightarrow \dot{\quat q} = \frac{1}{2}\quat q\quat \omega.
\label{eqn:quat-eom}
\end{equation}

The second equation of motion are Euler's equations of motion given our expression for tidal torque. Namely,
\begin{eqnarray}
I_1 \dot \omega_1 - \omega_2 \omega_3 (I_2 - I_3) &= \tau_1\\
I_2 \dot \omega_2 - \omega_3 \omega_1 (I_3 - I_1) &= \tau_2\\
I_3 \dot \omega_3 - \omega_1 \omega_2 (I_1 - I_2) &= \tau_3
\label{eqn:omega-eom}
\end{eqnarray}
The components of $\bm \tau$ are given by equation \ref{eqn:tidal-torque}, which was also in the body-fixed frame.

\subsection{Initial Values}
We assume that the asteroid begins the simulation in its minimum energy configuration; that is, with $\bm \omega \parallel \unit z$. The initial conditions are therefore that $\bm \omega = \omega \unit z$, and the quaternion is fixed to rotate $\hat z$ to $\bm \Omega$ in the inertial frame, which fixes Euler angles $\beta$ and $\alpha$. The third Euler angle $\gamma$ represents the initial roll of the asteroid, and therefore cannot be determined from physical data. However, it is an initial condition of the simulation. Once the Euler angles are known, the quaternion is computed via equation \ref{eqn:euler-to-quat}.

@@

\jtd{Running Jan 13}

@@

\bibliography{blank}{}
\bibliographystyle{aasjournal}
